Multiset-trie data structure

### 2023-02-20 tex/tex/s_math-Theorems.tex
all mathematical improvements

\begin{Lemma}\label{l:prob-node-existence}
Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
with probability
\begin{equation}
p_i=1-(1-p)^{n^{\sigma + 1 -i}}.
\end{equation}
\end{Lemma}
\begin{proof}
Let $v$ be an arbitrary node in a multiset-trie on an arbitrary level $i.$ Consider
the subtree with the root $v$ and call it $v$-subtree. Since the height of the
multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-subtree.
Taking into account that the root node has height 1, the height of the $v$-subtree is
%
\[
h_v = \sigma + 1 - i.
\]
%
A node in a multiset-trie exists if at least one node exists on the leaf level of its
subtree, i.e. a node on the level $\sigma + 1$ that belongs to $v$-subtree. The possible
number of nodes on the leaf level of $v$-subtree can be easily calculated knowing its height.
It is equal to
%
\[
n^{\sigma + 1 - i}.
\]
%
A node at level $\sigma+1$ exists with probability $p,$ where $p = P(m\in M).$
Thus, the probability that there are no nodes on leaf level in $v$-subtree is
%
\[
(1-p)^{n^{\sigma +1 - i}}.
\]
%
The claim follows by taking the complement probability of the above result.

%
\end{proof}

However, in order to determine the distribution of $\xi_{ij}$, one needs a lemma of a different kind.

\begin{Lemma}\label{l:prob-children}
Suppose that a node $v$ exists at level $1\leq i\leq\sigma$.
Then the number of its children $\xi_{iv}$ is modeled by a zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$. In particular,
the probability of node $v$ having $k$ children equals to
\begin{equation}\label{eq:pdf}
P(\xi_{iv} = k) = \frac{\binom{n}{k} (1-p_{i+1})^{n-k}}{1-(1-p_{i+1})^n}
\end{equation}
and the corresponding probability generating function equals to
\begin{equation}\label{eq:generating_func}
G_i(z) = \frac{(1+p_{i+1}(z-1))^n - (1-p_{i+1})^n}{1-(1-p_{i+1})^n}.
\end{equation}
\end{Lemma}
\begin{proof}
In order to prove the lemma, we have to show that
$\xi_{iv}\sim\mathcal{B}_0(n, p_{i+1}).$
Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
definition of the multiset-trie, a node exists at level $i$ if and only if
it has at least one child. Note that this is not true for the nodes on the leaf
level $\sigma + 1.$ Implies, a node on level $i$ can have
$k\in\{ 1,2,\ldots, n\}$ children. Let $X_0, X_1, \ldots, X_{n-1}$ be random
variables, they are defined as follows:
\[
X_k = \begin{cases}
0 & \textrm{child $k$ of node $v$ does not exist} \\
1 & \textrm{child $k$ of node $v$ exists} \\
\end{cases}
\]
As it was shown in previous Lemma~\ref{l:prob-children}, the distribution of 
$X_k$ is $X_k\sim Bernoulli(p_{i+1}).$ Since our model assumes that all the 
multisets in $M$ are chosen uniformly at random, the variables 
$X_k,X_l$ are independent for $k\neq l.$ But in our case the node $v$ can not 
have 0 children, so the sum $\sum_{k=1}^n X_k$ has a zero-truncated binomial 
distribution:
%
\[
\sum_{k=1}^n X_k \sim\mathcal{B}_0(n,p_{i+1})
\]
%
which completes the proof.

@@

\begin{Corollary}\label{cor:exp_nodes_param}
Let $M_1, M_2, \alpha_i$ and $\beta_i$ be defined as in previous Definition~\ref{def:params}, 
then the expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_1}$ that contains all the multisets 
from the set $M_1$ is equal to 
\begin{equation}\label{eq:nodes_sub_param}
\mathbb{E}( |\mathcal{M}_{M_1}| ) = \sum_{i=1}^{\sigma + 1} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}}.
\end{equation}
The expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_2}$ that contains all the multisets 
from the set $M_2$ is equal to 
\begin{equation}\label{eq:nodes_super_param}
\mathbb{E}( |\mathcal{M}_{M_2}| ) = \sum_{i=1}^{\sigma + 1} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}.
\end{equation}
\end{Corollary}
%
\begin{proof}
Using the results from Theorem~\ref{thm:exp_level} and Theorem~\ref{thm:exp_nodes} 
we derive the formulas~(\ref{eq:nodes_sub_param}) and~(\ref{eq:nodes_super_param}) 
by specifying the possible number of nodes on every level in the multiset-trie according 
to the multiset $m.$ Note that the formula~(\ref{eq:nodes_level}) assumes that on every 
level but the first one, there are $n$ possible nodes. Given sub-multiset or super-multiset 
query and an input multiset $m$ the number of nodes that will be traversed on level $i$ is 
defined by the number $k_{i-1}+1$ or $n-k_{i-1}-1$ for $i\geq 2.$ On level $i=1$ 
there is only one root node in any multiset-trie $\mathcal{M},$ which always exists 
if $M\neq\emptyset$ and is traversed for any type of query (sub-multiset and 
super-multiset).

@@

\subsubsection{SubmsetExistence and SupermsetExistence}\label{ss:exists}
We start the analysis of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} with an observation. Our theoretical model assumes 
that all the multisets are inserted into multiset-trie at random. It was already concluded 
that the probability distribution function $P(m\in M)$ has an impact on the size of 
multiset-trie $\mathcal{M}.$ Moreover, this distribution influences on the performance 
of the functions \textsc{submsetExistence} and \textsc{supermsetExistence} even more.

@@

\subsection{Space complexity}\label{s:spacecomplexity}
As in any efficient algorithm, there is always some trade-off between space and 
time complexity. While offering efficient sub- and super-multiset queries, an 
additional space must be provided for multisets storage. Clearly, the cardinality 
of the set $M$ is smaller than the size of $\mathcal{M},$ because the number 
of multisets in $\mathcal{M}$ is equal to the number of nodes only on the leaf 
level. The figure~\ref{f:exp-nodes} demonstrates the relation between the number 
of multisets stored and the number of nodes needed for storage, where parameters 
$\sigma$ and $n$ are $26$ and $10,$ respectively.

\begin{figure}[h!]
\center
\includegraphics[width=.4\textwidth, keepaspectratio]{exp-nodes-on-probab.pdf}
\caption{$\mathbb{E}(|\mathcal{M}|)$ and $\mathbb{E}(|M|)$ on probability.}
\label{f:exp-nodes}
\end{figure}

@@

\begin{figure}[h!]
\center
\includegraphics[width=.4\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob.pdf}
\caption{Ratio $\mathbb{E}(\frac{|\mathcal{M}|}{|M|})$ on $p.$}
\label{f:ratio-exp-msets}
\end{figure}

Note that analyzing the graph on figure~\ref{f:ratio-exp-msets} we can safely 
say that the upper bound for the ratio is $\sigma + 1.$ The argument holds, 
because of the limit 
\begin{equation}
\lim_{p\rightarrow 0^+} \mathbb{E}(\xi_i) = 1,
\end{equation}
where $\xi_i$ is the number of nodes on $i$-th level and $1\leq i \leq \sigma + 1.$

@@

Since the ratio $\sigma + 1$ can be obtained for a very specific case only and 
with a small increase in probability, the ratio drops rapidly it can be concluded 
that the space complexity of the multiset-trie is $O(|M|).$

>>>>

\begin{Lemma}\label{l:prob-node-existence}
Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
with probability
\begin{equation}
p_i=1-(1-p)^{n^{\sigma + 1 -i}}.
\end{equation}
\end{Lemma}
\begin{proof}
Let $v$ be an arbitrary node in a multiset-trie on an arbitrary level $i.$ Consider
the subtree with the root $v$ and call it $v$-subtree. Since the height of the
multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-subtree.
Taking into account that the root node has height 1, the height of the $v$-subtree is
%
\[
h_v = \sigma + 1 - i.
\]
%
A node in a multiset-trie exists if at least one node exists on the leaf level of its
subtree, i.e. a node on the level $\sigma + 1$ that belongs to $v$-subtree. The possible
number of nodes on the leaf level of $v$-subtree can be easily calculated knowing its height.
It is equal to
%
\[
n^{\sigma + 1 - i}.
\]
%
A node at level $\sigma+1$ exists with probability $p,$ where $p = P(m\in M).$
Thus, the probability that there are no nodes on leaf level in $v$-subtree is
%
\[
(1-p)^{n^{\sigma +1 - i}}.
\]
%
The claim follows by taking the complement probability of the above result. 
\end{proof}

Recall that for any given discrete random variable $X$ taking values over positive non-negative integral values, one can define 
a so-called \emph{probability generating function} (PGF, for short) $G_X$ to be a formal power series defined as 
\begin{align*}
G_X(z)=\sum_{i\ge 0} Pr(X=i)z^i.
\end{align*}
While the PGFs are usually not meant to be evaluated for concrete values of $z$, certain values have special interpretation when used in PGF, or in a derivation(s) of PGF. 
For instance, $Pr(X=0)=G_X(0)$, and also $\mathbb{E}(X)=G'_X(1)$.
For many values of $z$, the function $G_X$ may not converge to any finite value. For this reason it is a common notation to write, in particular
\[
G_X(1^-)=\lim_{z\nearrow 1}G_X(z).
\]

In what follows we will denote a Bernoulli-distributed random variable with parameter $p$ as $Bernoulli(p)$.
Furthermore we denote
zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$ by $\mathcal{B}_0(n,p_{i+1})$. Finally, $\overset{d}{=}$ indicates so-called equality in distribution.

\begin{Example}
The probability generating function of a binomial random variable $\mathcal B(n,p)$, the number of successes in $n$ trials, with probability $p$ of success in each trial, is

\begin{align*}
 G_{\mathcal B(n,p)}(z)=G(z)
 &=\sum_{k = 0}^n \binom n k p^k  (1 - p)^{n - k} z^k\\
 &= \sum_{k = 0}^n \binom n k  (p z)^k  (1 - p)^{n - k}\\
 &=\left((1-p)+pz\right)^{n}.
\end{align*}

It is thus easy to compute expectation
\[
\mathbb{E}(\mathcal{B}(n,p))=G'(1)=np.
\]

\end{Example}

We now focus our attention to the distribution of $\xi_{ij}$.

\begin{Lemma}\label{l:prob-children}
Suppose that a node $v$ exists at level $1\leq i\leq\sigma$.
Then the number of its children $\xi_{iv}$ is modeled by a zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$. In particular,
the probability of node $v$ having $k$ children equals to
\begin{equation}\label{eq:pdf}
P(\xi_{iv} = k) = \frac{\binom{n}{k} (1-p_{i+1})^{n-k}}{1-(1-p_{i+1})^n}
\end{equation}
and the corresponding probability generating function equals to
\begin{equation}\label{eq:generating_func}
\mathcal{B}_0(n,p_{i+1})\overset{d}{=}
G_i(z) = \frac{(1+p_{i+1}(z-1))^n - (1-p_{i+1})^n}{1-(1-p_{i+1})^n}.
\end{equation}

\end{Lemma}
\begin{proof}
In order to prove the lemma, we have to show that
$\xi_{iv}\sim\mathcal{B}_0(n, p_{i+1}).$
Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
definition of the multiset-trie, a node exists at level $i$ if and only if
it has at least one child. Note that this is not true for the nodes on the leaf
level $\sigma + 1.$ Implies, a node on level $i$ can have
$k\in\{ 1,2,\ldots, n\}$ children. Let $X_0, X_1, \ldots, X_{n-1}$ be random
variables, they are defined as follows:
\[
X_k = \begin{cases}
0 & \textrm{child $k$ of node $v$ does not exist} \\
1 & \textrm{child $k$ of node $v$ exists} \\
\end{cases}
\]
As it was shown in previous Lemma~\ref{l:prob-children}, the distribution of 
$X_k$ is $X_k\sim Bernoulli(p_{i+1}).$ Since our model assumes that all the 
multisets in $M$ are chosen uniformly at random, the variables 
$X_k,X_l$ are independent for $k\neq l.$ But in our case the node $v$ can not 
have 0 children, so the sum $\sum_{k=1}^n X_k$ has a zero-truncated binomial 
distribution:
%
\[
\sum_{k=1}^n X_k \overset{d}{=} \mathcal{B}_0(n,p_{i+1}),
\] 
%
which completes the proof.

@@

\begin{Corollary}\label{cor:exp_nodes_param}
Let $M_1, M_2, \alpha_i$ and $\beta_i$ be defined as in  Definition~\ref{def:params}, 
then the expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_1}$ that contains all the multisets 
from the set $M_1$ is equal to 
\begin{equation}\label{eq:nodes_sub_param}
\mathbb{E}( |\mathcal{M}_{M_1}| ) = \sum_{i=1}^{\sigma + 1} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}}.
\end{equation}
The expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_2}$ that contains all the multisets 
from the set $M_2$ is equal to 
\begin{equation}\label{eq:nodes_super_param}
\mathbb{E}( |\mathcal{M}_{M_2}| ) = \sum_{i=1}^{\sigma + 1} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}.
\end{equation}
\end{Corollary}
%
\begin{proof}
Using the results from Theorem~\ref{thm:exp_level} and Theorem~\ref{thm:exp_nodes} 
we derive the formulas~(\ref{eq:nodes_sub_param}) and~(\ref{eq:nodes_super_param}) 
by specifying the possible number of nodes on every level in the multiset-trie according 
to the multiset $m.$ Note that the formula~(\ref{eq:nodes_level}) assumes that on every 
level but the first one, there are $n$ possible nodes. Given sub-multiset or super-multiset 
query and an input multiset $m$ the number of nodes that will be traversed on level $i$ is 
defined by the number $k_{i-1}+1$ or $n-k_{i-1}-1$ for $i\geq 2.$ On level $i=1$ 
there is only one root node in any multiset-trie $\mathcal{M},$ which always exists 
if $M\neq\emptyset$ and is traversed for any type of query (sub-multiset and 
super-multiset).

@@

\subsubsection{SubmsetExistence and SupermsetExistence}\label{ss:exists}
We start the analysis of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} with an observation. Our theoretical model assumes 
that all the multisets are inserted into multiset-trie at random. It was already concluded 
that the probability distribution function $P(m\in M)$ has an impact on the size of 
multiset-trie $\mathcal{M}.$ 
Moreover, this distribution influences the performance 
of the functions \textsc{submsetExistence} and \textsc{supermsetExistence} even more.

@@

\subsection{Space complexity}\label{s:spacecomplexity}
As in any efficient algorithm, there is always some trade-off between space and 
time complexity. While offering efficient sub- and super-multiset queries, an 
additional space must be provided for multisets storage. 
%Clearly, the cardinality 
%of the set $M$ is smaller than the size of $\mathcal{M},$ because the number 
%of multisets in $\mathcal{M}$ is equal to the number of nodes only on the leaf 
%level. 
As we increase the parameter $p$, the number of sets we need to store (i.e. number of leaves) increases linearly. The figure~\ref{f:exp-nodes} shows that the additional overhead space that our data-structure uses does not increase as the tree is becoming more and more dense.

\begin{figure}[h!]
\center
\includegraphics[width=.4\textwidth, keepaspectratio]{exp-nodes-on-probab.pdf}
\caption{Size of a tree ($\mathbb{E}(|\mathcal{M}|)$) compared to the number of stored sets ($\mathbb{E}(|M|)$), as the probability parameter $p$ increases.
Parameters 
$\sigma$ and $n$ are $26$ and $10,$ respectively.
The additional overhead space that our data-structure uses does not increase as the tree is becoming more and more dense.
}
\label{f:exp-nodes}
\end{figure}

@@

\begin{figure}[h!]
\center
\includegraphics[width=.4\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob1.pdf}
\caption{The ratio of "data overhead" derived from Figure \ref{f:exp-nodes}, as the density $p$ of our data-structure increases.
Unless we are in a very sparse setting, the trade-off in terms of space is small.}
\label{f:ratio-exp-msets}
\end{figure}

Analyzing the graph on figure~\ref{f:ratio-exp-msets} we can safely 
say that the upper bound for the ratio is $\sigma + 1.$ The argument holds, 
because of the limit 
\begin{equation}
\lim_{p\rightarrow 0^+} \mathbb{E}(\xi_i) = 1,
\end{equation}
where $\xi_i$ is the number of nodes on $i$-th level and $1\leq i \leq \sigma + 1.$

@@

Since the ratio $\sigma + 1$ can be obtained for a very specific case only and 
with a small increase in probability, the ratio drops rapidly and it can be concluded 
that the space complexity of the multiset-trie is $O(|M|).$

\subsection{An informal comment on the mathematical analysis}
In the next section we show how our data-structure preforms in certain environments.
We will see, in our view, great results throughout various experiments, even when compared with state-of-the-art approaches like Inverted Index.

It was very frustrating to us that, despite those indications, it seems that this structure is very difficult to analyze. 
Even after restricting the analysis to the (usually simple) case when the input data is uniformly distributed, and after utilising reliable probabilistic tools, our approach unfortunately did not give as clear results as we hoped for. 
We would be very keen to learn of the approach or technique which would shed more insight into the mathematical analysis of our data-structure and its algorithms.
